---
title: "An Introduction to **ClassifyR**"
author: Dario Strbenac, Grahan Mann, Jean Yang, John Ormerod
institute: The University of Sydney, Australia
output: 
  BiocStyle::html_document:
    toc: true
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{An Introduction to the ClassifyR Package}
---

```{r, echo = FALSE, results = "asis"}
options(width = 130)
set.seed(44)
```

## Overview

**ClassifyR** provides a structured pipeline for cross-validated classification. Classification is viewed in terms of four stages, data transformation, feature selection, classifier training, and prediction. The stages can be run in any order that is sensible.

Each step can be provided with custom functions that follow some rules about parameters. The driver function *runTests* implements different varieties of cross-validation. They are:

* Ordinary k-fold cross-validation
* Permutation of the order of samples followed by k-fold cross-validation
* Repeated x% test set cross-validation
* leave-k-out cross-validation.

*runTests* can use parallel processing capabilities in R to speed up cross-validations when many CPUs are available. The output of *runTests* is a *ClassifyResult* object which can be directly used by the performance evaluation functions. The process of classification is summarised by a flowchart.


```{r, echo = FALSE}
htmltools::img(src = knitr::image_uri("ClassifyRprocedure.png"), 
               style = 'margin-left: auto;margin-right: auto')
```

Importantly, ClassifyR implements a number of methods for classification using different kinds of changes in measurements between classes. Most classifiers work with features where the means are different. In addition to changes in means, **ClassifyR** also allows for classification using differential deviation (changes in scale) and differential distribution (changes in location and/or scale).

In the following sections, some of the most useful functions provided in **ClassifyR** will be demonstrated. However, a user can provide any feature selection, training, or prediction function to the classification framework, as long as it meets some simple rules about the input and return parameters. See the last section of this guide "Rules for New Functions" for a description of these.

## Comparison to Existing Classification Frameworks

There are a few other frameworks for classification in R. The table below provides a comparison of which features they offer.

Package | Run User-defined Classifiers | Parallel Execution on any OS | Parameter Tuning | Intel DAAL Performance Metrics | Ranking and Selection Plots | Class Distribution Plot | Error Heatmap | Direct Support for MultiAssayExperiment Input
-------|--------|-------|--------|--------|---------|-----------|----------
**ClassifyR**    | Yes | Yes | Yes | Yes | Yes | Yes | Yes | Yes
caret        | Yes | Yes | Yes | No  | No  | No  | No  | No
MLInterfaces | Yes | No  | No  | No  | No  | No  | No  | No
MCRestimate  | Yes | No  | Yes | No  | No  | No  | No  | No
CMA          | No  | No  | Yes | No  | No  | No  | No  | No

## Case Study : Classifying Cancer Types

To demonstrate some key features of ClassifyR, a small dataset consisting of 35 features and 20 blood samples from stage 4 melanoma samples will be used to quickly obtain results. The 20 patients which were treated by either nivolumab or pembrolizumab are divided into 11 responders and 9 non-responders. The journal article corresponding to the dataset was published in 2018 and is titled [High-dimensional Single-cell Analysis Predicts Response to Anti-PD-1 Immunotherapy](https://www.nature.com/articles/nm.4466). Although thousands of cells were measured for each sample, those measurements have been summarised to 1 value per protein by taking the median of all cells' measurements for a particular protein.

```{r}
library(ClassifyR)
data(melanomaResponse) # Replace with real dataset soon. This is rnorm.
measurements[1:5, 1:5]
head(classes)
```

The numeric matrix variable *measurements* stores the normalised values of the protein abundances for each sample and the factor vector *classes* identifies which class the samples belong to.

For more complex datasets with multiple kinds of experiments (e.g. DNA methylation, copy number, gene expression on the same set of samples) a **MultiAssayExperiment** is recommended for data storage and supported by **ClassifyR**'s methods.

### runTests Driver Function of Cross-validated Classification

*runTests* is the main function in **ClassifyR** which handles the sample splitting and parallelisation, if used, of cross-validation. To begin with, a simple classifier will be specified. It uses a **limma** moderated t-test ranking for feature selection and DLDA for classification. The *limmaSelection* function also uses DLDA for estimating a resubstitution error rate for a number of top-*f* ranked features, as a heuristic for picking *f* features from the feature ranking which are used in the training and prediction stages of classification. This classifier relies on differences in means between classes.

```{r, tidy = FALSE}
resubstituteParams <- ResubstituteParams(nFeatures = 2:10,
                                         performanceType = "balanced error",
                                         better = "lower")
DMresults <- runTests(measurements, classes, "Melanoma", "Different Means",
                      validation = "leaveOut", leave = 1,
                      params = list(SelectParams(limmaSelection, "Moderated t Statistic",
                                                 resubstituteParams = resubstituteParams),
                                    TrainParams(DLDAtrainInterface, doesTests = FALSE),
                                    PredictParams(DLDApredictInterface,
                                                  getClasses = function(result)
                                                               result[["class"]])),
                      parallelParams = bpparam(), verbose = 1)
DMresults
```

Here, leave-one-out cross-validation (LOOCV) is specified by the values of *validation* and *leave*. The specification of feature selection and classifier algorithm is made by the list of parameter classes specified to *params*. Some classifiers available from repositories like CRAN do training and test set prediction with one function and others with two separate functions. Hence, it's necessary to specify a value for *doesTests* to the *TrainParams* constructor whenever doing classification. For computers with more than 1 CPU, the number of cores to use can be given to *runTests* by using the argument *parallelParams*. Although not used in this example, the parameter *seed* is important to set for result reproducibility when doing a cross-validation such as 100 sample permutations and 5 folds, because it employs randomisation to partition the samples into folds. For more details about *runTests* and the parameter classes used by it, consult the help pages of such functions.

### Evalulation of a Classification

The most frequently selected protein can be identified using the *distribution* function and its relative abundance values for all samples can be displayed visually by *plotFeatureClasses*.

```{r, fig.height = 12, fig.width = 12, results = "hold", message = FALSE}
selectionPercentages <- distribution(DEresults, plot = FALSE)
mostChosen <- names(sort(selectionPercentages, decreasing = TRUE)[1])
bestProteinPlot <- plotFeatureClasses(measurements, classes, mostChosen, dotBinWidth = 0.25)
```

The means of the protein levels are substantially different between the responder and non-responder patients. *plotFeatureClasses* can also plot categorical data, such as may be found in a clinical data table, as a bar chart.

Classification error rates, as well as many other prediction performance measures, can be calculated with *calcCVperformance*. Next, the balanced error rate is calculated considering all samples, each of which was in the test set once. The balanced error rate is defined as the average of the classification errors of each class.

```{r}
DEresults <- calcCVperformance(DEresults, "balanced error")
DEresults
performance(DEresults)
```

The error rate is small. If a cross-validation scheme such as 100 permutations, 5 folds was used instead, there would be 100 data sets to cross-validate and numerous predictions made of each sample, resulting in 100 values for each performance metric. Then, the function *performancePlot* could be used to visualise the distribution of the metric and even compare different classifiers. See the documentation of *performancePlot* for more details.

### Comparison of Different Classifications

The *samplesMetricMap* function allows the visual comparison of sample-wise error rate or accuracy measures from different *ClassifyResult* objects. Firstly, a classifier will be run that uses Kullback–Leibler divergence ranking and resubstitution error as a feature selection heursitic and a naive Bayes classifier for classification. This classification will use features that have either a change in location or in scale between classes.

```{r, tidy = FALSE}
DDresults <- runTests(measurements, classes, "Melanoma", "Differential Distribution",
                      validation = "leaveOut", leave = 1,
                      params = list(SelectParams(KullbackLeiblerSelection,
                                                 resubstituteParams = resubstituteParams),
                                    TrainParams(naiveBayesKernel, doesTests = TRUE),
                                    PredictParams(predictor = function(){}, getClasses = function(result) result,
                                                  weighted = "weighted", weight = "height difference",
                                                  returnType = "both")),
                                verbose = 1)
DDresults
```

Here, *doesTests* is TRUE because *naiveBayesKernel* has input parameters for the traing data and the test data. The naive Bayes kernel classifier has many options specifying how the distances between class densities are used. For more information, consult the documentation of the function named *naiveBayesKernel*.

Now, the classification error for each sample is calculated for both the differential means and differential distribution classifiers.

```{r, fig.width = 10, fig.height = 7}
library(grid)
DMresults <- calcCVperformance(DEresults, "sample error")
DDresults <- calcCVperformance(DDresults, "sample error")
resultsList <- list(Abundance = DMresults, Distribution = DDresults)
errorPlot <- samplesMetricMap(resultsList, metric = "error")
grid.draw(errorPlot)
```

```{r, fig.width = 10, fig.height = 7}
accuracyPlot <- samplesMetricMap(resultsList, metric = "accuracy")
```

The *performancePlot* function allows the comparison of overall performance measures, such as accuracy and error rate.

```{r, fig.width = 6.8, fig.height = 5}
errorBoxes <- performancePlot(list(DEresults, DVresults, DDresults[["weight=crossover distance"]]),
                              performanceName = "Balanced Error Rate",
                              boxFillColouring  = "None", boxLineColouring = "None",
                              title = "Errors Across Classification Types")
```

This plots the balanced error rates of the three kinds of classification together.

## Using an Independent Test Set

Sometimes, cross-validation is unnecessary. This happens when studies have large sample sizes and are well-designed such that a large number of samples is prespecified to form a test set. The classifier is only trained on the training sample set, and makes predictions only on the test set.

To demonstrate how this kind of analysis can be done with ClassifyR, a training and a test dataset are simulated, with 500 features and 50 samples in each set. 25 features will be differentially expressed in the training set. 25 features will be differentially expressed in the test set, and 15 of them will be the same features as in the training set.

```{r}
trainingExpr <- matrix(rnorm(500 * 50, 9, 3), ncol = 50)
trainingClasses <- factor(rep(c("Healthy", "Diseased"), each = 25), levels = c("Healthy", "Diseased"))
trainingExpr[101:125, trainingClasses == "Diseased"] <- trainingExpr[101:125, trainingClasses == "Diseased"] - 2

testingExpr <- matrix(rnorm(500 * 50, 9, 3), ncol = 50)
testingClasses <- factor(rep(c("Healthy", "Diseased"), each = 25), levels = c("Healthy", "Diseased"))
testingExpr[111:135, testingClasses == "Diseased"] <- testingExpr[111:135, testingClasses == "Diseased"] - 2
```

There are two matrices; one for the training set and one for the test set. Since *runTest* expects a single expression object, the matrices are combined putting the columns together and concatenating the classes vectors. The specification of training and testing samples happens by providing the column numbers of each group of samples.

```{r}
allExpr <- cbind(trainingExpr, testingExpr)
allClasses <- unlist(list(trainingClasses, testingClasses))
independentResult  <- runTest(allExpr, allClasses, datasetName = "Simulation", classificationName = "DE",
                              training = 1:50, testing = 51:100)
independentResult
```

## Cross-validating Selected Features on a Different Dataset

Once a cross-validation classification is complete, the usefulness of the features selected may be explored in another dataset. *previousSelection* is a function which takes an existing ClassifyResult object and returns the features selected at the equivalent iteration which is currently being processed. This is necessary, because the models trained on one dataset are not directly transferrable to a new dataset. The classifier training is redone.

## Generating a ROC Plot

Some classifiers can be set to output scores or probabilities representing how likely a sample is to be from one of the classes, rather than class labels. This enables different score thresholds to be tried, to generate pairs of false positive and false negative rates. The naive Bayes classifier and Fisher discriminant analysis used previously had the *returnType* variable set to *"both"*, so labels and scores were both stored in the classification result. Setting *returnType* to *"score"* is also sufficient. Many existing classifiers in other R package also have an option that allows a score or probability to be calculated.

```{r, fig.height = 5, fig.width = 6}
ROCcurves <- ROCplot(list(DVresults, DDresults[["weight=crossover distance"]]))
```

## Parameter Tuning

Some classifiers allow the setting of a tuning parameter, which controls some aspect of their model learning. An example of doing parameter tuning with a linear SVM is presented. The SVM has a single tuning parameter, the cost. Higher values of this parameter penalise misclassifications more.

This is acheived in ClassifyR by providing a variable called *tuneParams* to the TrainParams container constructor. *tuneParams* is a named list, with the names being the names of the tuning variables, and the contents being vectors of values to try. If *tuneParams* has more than one element, all combination of values of the tuning variables are tried. The performance criterion specified to *resubstituteParams* is also used as the criterion for choosing the best tuning parameter(s). This means that any of the performance measures calculated by **ROCR** can be used.

A linear SVM is demonstrated. It only has one tuning parameter, the cost value.

```{r}
library(e1071) # Provides SVM functions.
resubstituteParams = ResubstituteParams(nFeatures = c(25, 50, 75, seq(100, 1000, 100)),
                                        performanceType = "balanced", better = "lower")
SVMresults <- runTests(ovarSet, "Ovarian Cancer", "Differential Expression", validation = "permute",
                       permutations = 5, folds = 3,
                       params = list(SelectParams(limmaSelection, resubstituteParams = resubstituteParams),
                                     TrainParams(svm, TRUE, doesTests = FALSE, kernel = "linear",                                                             resubstituteParams = resubstituteParams,
                                                 tuneParams = list(cost = c(0.01, 0.1, 1, 10))),
                                     PredictParams(predict, TRUE, getClasses = function(result) result)),
                       parallelParams = bpparam(), verbose = 1)
```

The chosen values of the parameters are stored for every validation, and can be accessed with the *tunedParameters* function.

```{r}
length(tunedParameters(SVMresults))
tunedParameters(SVMresults)[[1]]
```

These are the cost values chosen for the three folds of the first resampling.

## Conclusion

When many replicates per class are available, differential variability or distribution classification may have better prediction perfomance than traditional differential expression analysis. Judging by feature selection, the probes chosen for their differential distribution have much stronger differences than those for expression.

## Rules Regarding Input Variables of New Functions

The required inputs and type of output that each stage of classifiation has is summarised by the table below. The functions can have any number of other arguments after the set of arguments which are mandatory.

```{r, echo = FALSE}
htmltools::img(src = knitr::image_uri("functionRules.png"), 
               style = 'margin-left: auto;margin-right: auto')
```

The argument *verbose* is sent from *runTest* to these functions so they must handle it. In the ClassifyR framework, *verbose* is a number which indicates the amount of progress messages to be printed. If verbose is 0, no progress messages are printed. If it is 1, only one message is printed for every 10 cross-validations completed. If it is 2, in addition to the messages printed when it is 1, a message is printed each time one of the stages of classification (transformation, feature selection, training, prediction) is done. If it is 3, in addition to the messages printed for values 1 and 2, progress messages are printed from within the classification functions themselves.

A version of each inlcuded transformation, selection, training and prediction function is typically implmented for (1) a numeric matrix for which the rows are for features and columns are for samples and a factor vector of the same length as the number of columns of the matrix, (2) a *DataFrame* and a class specification and (3) a *MultiAssayExperiment* which stores class information in the *colData* slot's *DataFrame* with column name "class". For the inputs which are not *DataFrame*, they are converted to one, because the other data types can be stored as a *DataFrame* without loss of information and the function which accepts a *DataFrame* contains the code to do the actual computations. At a minimum, a new function must have a method taking a *DataFrame* as input with the sample classses either stored in a column named "class" or provided as a factor vector. Although not required, providing a version of a function that accepts a numeric matrix with an accompanying factor vector and another version that accepts a MultiAssayExperiment is desirable to provide felxibility regarding input data. See the code of existing functions in the package for examples of this, if intending to contribute novel classification-related methods to **ClassifyR**.

## References
Strbenac D., Yang, J., Mann, G.J. and Ormerod, J. T. (2015) [ClassifyR: an R package for performance assessment of classification with applications to transcriptomics](http://bioinformatics.oxfordjournals.org/content/31/11/1851), *Bioinformatics*, 31(11):1851-1853
Strbenac D., Mann, G.J., Yang, J. and Ormerod, J. T. (2016) [Differential distribution improves gene selection stability and has competitive classification performance for patient survival](http://nar.oxfordjournals.org/content/44/13/e119), *Nucleic Acids Research*, 44(13):e119